---
title: "Lab 11"
author: "Catherine Diep"
format: html
---

> Q14: Generate a boxplot with a box per genotype, what could you infer from the relative
expression value between A/A and G/G displayed in this plot? Does the SNP effect the
expression of ORMDL3?

```{r}
data11 <- read.table("data11.txt")
```

```{r}
boxplot11 <- boxplot(exp~geno, data= data11, xlab="Genotype", ylab="Expression", main="SNP Effect on ORMDL3 Expression", notch=TRUE, varwidth=TRUE, col=c("red", "green", "blue"))
```
A/A is more highly expressed than G/G. The SNP does affect the expression of ORMDL3. 

> Q13: Read this file into R and determine the sample size for each genotype and their
corresponding median expression levels for each of these genotypes. 

```{r}
summary(boxplot11)
```
```{r}
boxplot11
```

```{r}
#subset(data11, geno%in% c("A/A"))
```
```{r}
#subset(data11, geno%in% c("A/G"))
```
```{r}
#subset(data11, geno%in% c("G/G"))
```


The median expression level for genotype A/A is 31.25, for A/G is 25.06, and for G/G is 20.07.The sample sizes are 108 for A/A, 233 for A/G, and 121 for G/G. 

```

