---
title: "Lab 17 - COVID Mini Project"
author: "Catherine Diep"
format: html
---

## Lab 17 - COVID-19 Vaccination Mini Project

##Getting Started

```{r}
# Import vaccination data
vax <- read.csv("covidzip.csv")
head(vax)
tail(vax)
```

> Q1 What column details the total number of people fully vaccinated?
Column 9 = persons fully vaccinated
> Q2 What column details the Zip code tabulation area?
Column 2 = ZIP code tabulation area
> Q3 What is the earliest date in this dataset? 
Earliest date = 2021-01-05
> Q4 What is the latest date in this dataset?
Latest date = 2023-03-07

```{r}
skimr::skim(vax)
```

>Q5. How many numeric columns are in this dataset? 
13  
>Q6. Note that there are “missing values” in the dataset. How many NA values there in the persons_fully_vaccinated column? 
20,965
>Q7. What percent of persons_fully_vaccinated values are missing (to 2 significant figures)? 
10.5%

```{r}
library(lubridate)
```
```{r}
today()
```
```{r}
# Specify that we are using the year-month-day format
vax$as_of_date <- ymd(vax$as_of_date)
```
```{r}
# How many days have passed since the the 1st vaccination in this dataset?
today() - vax$as_of_date[1]
```
```{r}
# How many days the dataset spans 
vax$as_of_date[nrow(vax)] - vax$as_of_date[1]
```

>Q9. How many days have passed since the last update of the dataset?

```{r}
today() - vax$as_of_date[139356]
```

>Q10. How many unique dates are there in the dataset? 

```{r}
length(unique(vax$as_of_date))
```

## Working with Zip Codes 
```{r}
library(zipcodeR)
geocode_zip('92037')
```
```{r}
zip_distance('92037','92109')
```
```{r}
# Pull census data from zip code 
reverse_zipcode(c('92037', "92109") )
```
```{r}
# Pull data for all ZIP codes in the dataset
#zipdata <- reverse_zipcode( vax$zip_code_tabulation_area )
```

## Focus on the San Diego ARea 
```{r}
# Subset to San Diego county only areas
library(dplyr)

sd <- filter(vax, county == "San Diego")

nrow(sd)
```

>Q11. How many distinct zip codes are listed for San Diego County?

```{r}
length(unique(sd$zip_code_tabulation_area))
```


>Q12. What San Diego County Zip code area has the largest 12 + Population in this dataset?
92154

```{r}
sd.12 <- filter(vax, county == "San Diego" &
                age12_plus_population > 10000)
which.max(sd.12$age12_plus_population)
```
```{r}
sd.12[49,]
```

>Q13. What is the overall average “Percent of Population Fully Vaccinated” value for all San Diego “County” as of “2022-11-15”? 
74%

```{r}
library(tidyr)
vaxna <- drop_na(vax,percent_of_population_fully_vaccinated)
percent <- filter(vaxna, county == "San Diego" &
                as_of_date == "2022-11-15")
mean(percent$percent_of_population_fully_vaccinated)
```

>Q14. Using either ggplot or base R graphics make a summary figure that shows the distribution of Percent of Population Fully Vaccinated values as of “2022-11-15”?

```{r}
library(ggplot2)
ggplot(vaxna, aes(percent_of_population_fully_vaccinated)) + geom_histogram()
```

## Focus on UCSD/La Jolla 

```{r}
ucsd <- filter(sd, zip_code_tabulation_area=="92037")
ucsd[1,]$age5_plus_population
```

>Q15. Using ggplot make a graph of the vaccination rate time course for the 92037 ZIP code area:

```{r}
ggplot(ucsd) +
  aes(as_of_date,
      percent_of_population_fully_vaccinated) +
  geom_point() +
  geom_line(group=1) +
  ylim(c(0,1)) +
  labs(x="Date", y="Percent Vaccinated")
```

## Comparing to similar sized areas 
```{r}
# Subset to all CA areas with a population as large as 92037
vax.36 <- filter(vax, age5_plus_population > 36144 &
                as_of_date == "2022-11-15")

head(vax.36)
```
>Q16. Calculate the mean “Percent of Population Fully Vaccinated” for ZIP code areas with a population as large as 92037 (La Jolla) as_of_date “2022-11-15”. Add this as a straight horizontal line to your plot from above with the geom_hline() function?

```{r}
mean(vax.36$percent_of_population_fully_vaccinated)
```
```{r}
ggplot(ucsd) +
  aes(as_of_date,
      percent_of_population_fully_vaccinated) +
  geom_point() +
  geom_line(group=1) +
  ylim(c(0,1)) +
  geom_hline(aes(yintercept=0.7190967))
  labs(x="Date", y="Percent Vaccinated")
```

>Q17. What is the 6 number summary (Min, 1st Qu., Median, Mean, 3rd Qu., and Max) of the “Percent of Population Fully Vaccinated” values for ZIP code areas with a population as large as 92037 (La Jolla) as_of_date “2022-11-15”?
0.38, 0.64, 0.72, 0.72, 0.79, 1.0

```{r}
fivenum(vax.36$percent_of_population_fully_vaccinated)
mean(vax.36$percent_of_population_fully_vaccinated)
```

Q18. Using ggplot generate a histogram of this data.
```{r}
ggplot(vax.36, aes(percent_of_population_fully_vaccinated)) + geom_histogram()
```
>Q19. Is the 92109 and 92040 ZIP code areas above or below the average value you calculated for all these above?
Below 

```{r}
vax %>% filter(as_of_date == "2022-11-15") %>%  
  filter(zip_code_tabulation_area=="92040") %>%
  select(percent_of_population_fully_vaccinated)
```
```{r}
vax %>% filter(as_of_date == "2022-11-15") %>%  
  filter(zip_code_tabulation_area=="92109") %>%
  select(percent_of_population_fully_vaccinated)
```
>Q20. Finally make a time course plot of vaccination progress for all areas in the full dataset with a age5_plus_population > 36144.

```{r}
vax.36.all <- filter(vax, age5_plus_population > 36144)


ggplot(vax.36.all) +
  aes(as_of_date,
      percent_of_population_fully_vaccinated, 
      group=zip_code_tabulation_area) +
  geom_line(alpha=0.2, color="blue") +
  ylim(c(0,1)) +
  labs(x="Date", y="Percent Vaccinated",
       title="Vaccination rate across California",
       subtitle="Only areas with a population above 36k are shown") +
  geom_hline(yintercept = mean(vax.36.all$percent_of_population_fully_vaccinated), linetype="dashed")
```

