---
title: "Class 05: Data Visualization with GGPLOT"
author: "Catherine Diep"
format: html
---

# Our first ggplot

To use the ggplot2 package, I first need to have it installed on my computer.

To install any package, we use the 'install.packages()' command.

Now can I use it? No - first we need to call 'library(ggplot2)'.

```{r}
library(ggplot2)
ggplot()
```

```{r}
mpg
```

Our first plot of displ vs hwy All ggplot() graphs are made in the same way - data + aes + geoms

```{r}
ggplot(mpg) +
  aes(x=displ, y=hwy) +
  geom_point()+
  geom_smooth()

```

I can add more layers:

```{r}
ggplot(mpg) +
  aes(x=displ, y=hwy) +
  geom_point()+
  geom_smooth(method = lm, se = FALSE)
```

# Plot of gene expression data

First, read this file from online.

```{r}
url <- "https://bioboot.github.io/bimm143_S20/class-material/up_down_expression.txt"
genes <- read.delim(url)
head(genes)
```

> Q. How many genes are in this dataset?

```{r}
nrow(genes)
```

What are the colnames?

```{r}
colnames(genes)
```

A first version plot of this data Condition 1 vs Condition 2

```{r}
ggplot(genes) +
  aes(x=Condition1, y=Condition2) +
  geom_point()
```

Let's add some color. To do this we can add the mapping of the color to the 'State' column in our data. 
```{r}
p <- ggplot(genes) +
  aes(x=Condition1, y=Condition2, col=State) +
  geom_point() 
```

> Q. How many genes are up regulated and down regulated?

```{r}
head(genes)
table(genes$State)
```

```{r}
p + scale_colour_manual( values=c("blue","gray","red") )+ labs(title= "Gene Expression Changes Upon Drug Treatment", x= "Control (No Drug)", y = "Drug Treatment")
```

