---
title: "Lab 10"
author: Catherine Diep
format: html
---

# Importing Candy Data
```{r}
candy_file <- read.csv("candy-data.csv")
candy = data.frame(candy_file, row.names=1)
head(candy)
```

>Q1. How many different candy types are in this dataset?
85 

```{r}
dim(candy_file)
```

>Q2. How many fruity candy types are in the dataset?
39 

```{r}
sum(candy_file$fruity, 1)
```

# What is your favorite candy?

```{r}
candy["Twix", ]$winpercent
```

>Q3. What is your favorite candy in the dataset and what is it’s winpercent value?
Caramel apple pop,s 34.5%

```{r}
candy["Caramel Apple Pops", ]$winpercent
```

>Q4. What is the winpercent value for “Kit Kat”?

```{r}
candy["Kit Kat", ]$winpercent
```

>Q5. What is the winpercent value for “Tootsie Roll Snack Bars”?

```{r}
candy["Tootsie Roll Snack Bars", ]$winpercent
```

Installing the skimr package to get a quick overview of a given dataset. 

```{r}
#install.packages("skimr")
library("skimr")
skim(candy)
```

>Q6. Is there any variable/column that looks to be on a different scale to the majority of the other columns in the dataset?
The histogram variable

>Q7. What do you think a zero and one represent for the candy$chocolate column?
If a candy is chocolate candy, the value will be 1; if not, it will be 0. 

>Q8. Plot a histogram of winpercent values

```{r}
hist(candy$winpercent)
```

>Q9. Is the distribution of winpercent values symmetrical?
The distribution is about symmetrical; it skews right. 

>Q10. Is the center of the distribution above or below 50%?
Below 

>Q11. On average is chocolate candy higher or lower ranked than fruit candy?
Higher

```{r}
mean(candy$winpercent[as.logical(candy$chocolate)])
mean(candy$winpercent[as.logical(candy$fruity)])
```

>Q12. Is this difference statistically significant?
Yes

```{r}
t.test(candy$fruity, candy$chocolate)
```

# Overall Candy Rankings

>Q13. What are the five least liked candy types in this set?
Nik L Nip, Boston Baked Beans, Chiclets, Super Bubble, Jawbusters

```{r}
head(candy[order(candy$winpercent),], n=5)
```


```{r}
library(dplyr)
candy %>% arrange(winpercent) %>% head(5)
```


>Q14. What are the top 5 all time favorite candy types out of this set?
Snickers, Kit Kat, Twix, Reese's Miniatures, Reese's Peanut Butter Cups

```{r}
candy %>% arrange(winpercent) %>% tail(5)
```

>Q15. Make a first barplot of candy ranking based on winpercent values.

```{r}
library(ggplot2)
ggplot(candy) + aes(winpercent, rownames(candy)) + geom_col()
```

>Q16. This is quite ugly, use the reorder() function to get the bars sorted by winpercent?

```{r}
library(ggplot2)
ggplot(candy) + aes(winpercent, reorder(rownames(candy),winpercent)) + geom_col()
```

```{r}
#Adding Colors
my_cols=rep("black", nrow(candy))
my_cols[as.logical(candy$chocolate)] = "chocolate"
my_cols[as.logical(candy$bar)] = "brown"
my_cols[as.logical(candy$fruity)] = "pink"

ggplot(candy) + 
  aes(winpercent, reorder(rownames(candy),winpercent)) +
  geom_col(fill=my_cols) 
```
>Q17. What is the worst ranked chocolate candy?
Payday

>Q18. What is the best ranked fruity candy?
Welchs fruit snacks 

#Taking a look at pricepercent

```{r}
library(ggrepel)

# How about a plot of price vs win
ggplot(candy) +
  aes(winpercent, pricepercent, label=rownames(candy)) +
  geom_point(col=my_cols) + 
  geom_text_repel(col=my_cols, size=3.3, max.overlaps = 5)
```

>Q19. Which candy type is the highest ranked in terms of winpercent for the least money - i.e. offers the most bang for your buck?
Tootsie Roll Midgies

```{r}
ord <- order(candy$pricepercent, decreasing = FALSE)
head( candy[ord,c(11,12)], n=5 )
```

>Q20. What are the top 5 most expensive candy types in the dataset and of these which is the least popular?
Nik L Nip

```{r}
ord <- order(candy$pricepercent, decreasing = TRUE)
head( candy[ord,c(11,12)], n=5 )
```

# Exploring the Correlation Structure 

```{r}
#install.packages("corrplot")
library(corrplot)
cij <- cor(candy)
corrplot(cij)
```

>Q22. Examining this plot what two variables are anti-correlated (i.e. have minus values)?
Chocolate and fruity 

>Q23. Similarly, what two variables are most positively correlated?
Chocolate and winpercent

# Principal Component Analysis

```{r}
pca <- prcomp(candy, scale=TRUE)
summary(pca)
```

```{r}
plot(pca$x[, 1:2])
```

```{r}
plot(pca$x[,1:2], col=my_cols, pch=16)
```

```{r}
# Make a new data-frame with our PCA results and candy data
my_data <- cbind(candy, pca$x[,1:3])

p <- ggplot(my_data) + 
        aes(x=PC1, y=PC2, 
            size=winpercent/100,  
            text=rownames(my_data),
            label=rownames(my_data)) +
        geom_point(col=my_cols)

p
```

```{r}
library(ggrepel)

p + geom_text_repel(size=3.3, col=my_cols, max.overlaps = 7)  + 
  theme(legend.position = "none") +
  labs(title="Halloween Candy PCA Space",
       subtitle="Colored by type: chocolate bar (dark brown), chocolate other (light brown), fruity (red), other (black)",
       caption="Data from 538")
```

Using Plotly for an interactive plot 

```{r}
library(plotly)
ggplotly(p)
```

```{r}
par(mar=c(8,4,2,2))
barplot(pca$rotation[,1], las=2, ylab="PC1 Contribution")
```

>Q24. What original variables are picked up strongly by PC1 in the positive direction? Do these make sense to you?
Fruity, hard, and pluribus - these values are like "outliers" so they have a greater effect on PC1. 